Stability and Change in Personal Culture

### 2019-12-06 manuscript9.Rmd
kevin's revisions

---
title: "Measuring Stability and Change in Personal Culture Using Panel Data"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-01.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to major theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???). Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time. Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller]. Attitudes can also be in a process of active updating but relatively easy to report, such as when a voter changes her registered partisan affiliation.

In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for estimating the prevalence of active updating using panel data. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By focusing on the pattern of attitude change at the individual level, we can describe the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary U.S. society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent responses and a few appear to elicit random responses. At the same time, people are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

These findings are generally somewhat more consistent with the settled dispositions model than with the active updating model. This is, they offer broad support to theories claiming that cultural change tends to come through generational turnover rather than persuasion. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification that to most specific policy positions. We argue that there is a place for both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.

@@

These models may seem similar at first glance, but they have very different implications for the pattern of individuals' responses over time. Equation \@ref(eq:AR) represents an _active updating_ model and Equation \@ref(eq:FE) represents a _settled dispositions_ model, as we explain below. Figure \@ref(fig:dag) shows these models in graphical form, which helps highlight their differences. In the next two sections, we consider each of these models and briefly discuss their connections to sociological theory. In the section after that, we discuss the conditions under which each model is more likely to apply.

@@

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from even earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property, and this active updating model generally, is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings]. Concerns about panel conditioning, in which a person's participation in a survey influences her subsequent answer choices, either by changing her non-survey behavior or by altering how they respond to the survey without changing behavior, also imply some sort of active updating on the part of the respondent [@warrenhalpernmanners].

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

@@

@@

Equation \@ref(eq:FE) represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are thus temporary influences only.

The SDM is most consistent with the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp. 66ff]. In other words, updating happened, but it happened in the past, prior to the time of the study. In this sense, $U_i$ reflects the "habitus." We will say more about this below.

@@

Metaphorically, one can think of the active updating model as a "traveler" and the settled dispositions model as a "visitor." That is, in the active updating model, the traveler wakes up in the morning, walks in a direction for a while, goes to sleep, and then wakes up the next day to pick up the journey exactly where she left off. In the settled dispositions model, the visitor also travels in a particular direction each day but goes home at the end of the day and wakes up in the same bed each morning.

@@

## Consistency in Response
Neither model makes any assumptions about the magnitude of of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change. Knowing that there is more wave-to-wave change in one variable than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.

## Combining the Models

These two dimensions of attitude change give us a two-by-two matrix of patterns for attitude change. The first dimension reflects what we call active updating, or the idea that more recent responses are better predictors of current responses that older responses. In other words, active updating means that when changes happen, they are more likely to persist than revert. We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses. People might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their prior positions. The second dimension reflects what we call stability, or how consistent people are in reporting their beliefs once we account for the amount of active updating.

With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them. Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix. We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes. Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.

\begin{center}
\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
\hline
 & Low Active Updating & High Active Updating \\ 
\hline
High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\ 
\hline
Low Consistency &  \\ 
\hline
\end{tabular}
\end{center}

### Age and Attitude Formation

In broad terms, cohort replacement theories of public opinion change posit that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo]. Implied in these models is the notion that individuals have a period of "cohortization" in which they form these initial beliefs.

To continue the metaphor above, cohort replacement theories imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition). The concept of a "cohort effect" is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman]. Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative [@vaiseylizardo]. To put it slightly differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick]. Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].

Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender. Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts. Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].

As Table ## suggests, items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults in the population, but high active updating for young people. Because these approaches generally assume that beliefs are prominent in the minds of respondents, the implication is that these beliefs are reported with high consistency in adulthood. At younger ages, however, we expect more volatility as attitude development occurs. We note, however, that there is some evidence that volatility of settled beliefs increases as people change as well [@krosnickalwin], but we do not expect this to be the general pattern.

### Inconsistent Opinions

Not all cultural change occurs through cohort replacement processes, and there is abundant evidence of large-scale public opinion shifts that occur in a span of time too short to be driven entirely by cohort replacement. This suggests that, at times, even middle-aged individuals update their opinions in persisting ways.

In contrast to cohort replacement theories, a range of theories posit that, outside of a well informed elite, individuals lack consistent opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have [@zaller; @converse]. For example, when asked whether the government should spend more on health care, forming an opinion requires someone to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result. Most people do not have this information readily stored in their minds.

Because of the cognitive demands of holding consistent opinions on these issues, these theories posit that individuals simply generate responses in the interview context, drawing on resources from their immediate social environment. This explains why survey respondents are susceptible to question-order, priming, and recency effects [@zaller]. However, these theories suggest that when questions of public policy receive significant media attention, individuals obtain more information and can more clearly articulate their positions on these issues. Shifts in public opinion can also come about when elites shift their opinions on issues, providing new signals to the general public.

In general, then, we expect low-profile and complex public policy issues to show a low rate of active updating and a low rate of consistency. By definition, most questions of public policy should fall into this group, since only a few issues can sustain a relatively high level of public attention. The small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do spotlight one issue that saw significant change in elite opinion. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, then, we might expect that public opinion shifts notably. This should manifest as a high degree of active updating for the whole population, not just young people.

Outside of a lack of information on public policy issues, there are two other reasons why individuals might demonstrate low consistency and low active updating. First, individuals might be unwilling to report socially undesirable beliefs if they hold them, and as a result obfuscate their true position. Similarly, some issues might not require a lot of information to comprehend, but might be so far from the public consciousness as to be nonsensical to respondents. We expect both of these conditions to produce a similar pattern of low active updating and low consistency.

### Updating Opinions

The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, people update their opinions across their lives in response to gathering new information and interacting with changing social environments.

Some of these approaches, common in political science, suggest that people consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman]. These models do not tend to make distinctions about which kinds of attitudes shift over time and which do not. However, for these models to be true, we would have to observe a high rate of active updating and a high rate of consistency.

Other theories suggest that only some beliefs are open to updating during adulthood. A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein]. These approaches include Lakoff’s [-@lakoff] model of moral politics, where views of the family serve as a template onto which to transpose questions of public policy, and Converse’s [-@converse] notion of social constraint, where political identity is used to form more specific policy stances as a result of information streams. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find little support for Lakoff's model and significant support for ideological identification as this "core" belief in the political space. If these individuals have the ability to regularly assess new information against these core beliefs, they should also be able to recognize their position in a multiple-choice survey questionnaire and report them with limited error [@vaisey09]. In other words, we expect that core beliefs such as ideological identification should have high consistency and low active updating, and conversely that items that show these traits to have the potential to be core beliefs.

In a similar vein of research focused on political polarization, Baldassari and Gelman [@baldassarigelman] find little increase in the correlation between specific policy positions over time, suggesting limited active updating of these items. What they do find is an increased alignment between partisan identification and policy positions. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, partisan affiliation should demonstrate the highest level of active updating, and that it should be reported with a high degree of consistency.

### Public vs. Private Culture

A final dimension that likely influences the consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances, which might lead to a low degree of consistency and active updating. However, when public culture provides a strong "scaffold" for individuals through clear signals of how their identities and social locations should influence how they should respond to a survey question, it becomes easier for them to maintain consistency [@lizardostrand].

We have touched on this idea already with our discussion of the idea that high-profile elite opinion change might drive opinion change in the broader public, or by noting that high-profile issues might be easier to report consistently because individuals receive signals that support that position. Along the same line, items that tap some public dimension of individual behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only "in a person's head" might demonstrate both higher rates of active updating and greater consistency.

### Toward Theoretical Synthesis

The clearest contrast in existing theories is between models that assume that beliefs are relatively settled during adulthood (cohort replacement theories), which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives (most other models), which would produce a pattern of high active updating. This disagreement thus forms the primary structure of our analysis. A lack of evidence in favor of active updating for most individuals in the population, coupled with evidence of active updating in younger members of the population and a high degree of consistency, would generally suggest that individuals form opinions early in life and maintain them throughout adulthood, consistent with cohort replacement and habitus theories of attitude development. We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating. However we should still observe high consistency and low active updating.

If what we find is a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions. Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work. We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

@@

Third, _are there systematic differences in item content between questions that fall in different parts of our matrix?_ The preceding sections have made some specific predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs. As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.

@@

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response. If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.

@@

That is, the SDM predicts that the average response of previous waves is the best predictor of the next response. If change is non-persisting (i.e., if our metaphorical visitor eventually always returns home), then taking the average of the last two places we saw her will be our best guess about the location of her home base.

@@

>>>>

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-01.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).

In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data. In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By classifying the pattern of change at the individual level, we can clarify when different accounts of aggregate change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S. society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that large numbers of U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent or random responses. At the same time, people are consistent on both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

Our findings offer broad support to theories claiming that cultural change comes through generational turnover rather than persuasion and social influence. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to specific policy positions. We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.

@@

These models may seem similar at first glance, but they have different implications for the pattern of individuals' responses we would observe over time. Equation \@ref(eq:AR) represents an _active updating_ model and Equation \@ref(eq:FE) represents a _settled dispositions_ model, as we explain below. Figure \@ref(fig:dag) shows these models in graphical form, which helps highlight their differences. In the next two sections, we consider each of these models and briefly discuss their connections to sociological theories.

@@

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

The concept of active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors. When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model. When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time. Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.

@@

A number of theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the population is becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller]. In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.

@@

Equation \@ref(eq:FE) represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are thus temporary influences only. At its core, this model reflects the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp. 66ff]. In other words, updating happened, but it happened in the past, prior to the time of the study. In this sense, $U_i$ reflects the "habitus."

The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007]. The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level. These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions (Mannheim 1952; Ryder 1965). Vaisey and Lizardo (2016), looking at a range of opinion items in the GSS, suggest that change for most likely comes in the form of cohort replacement.

On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this does not need to be the case. The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have (Zaller 1992; Converse 1964), and suggest that peoples’ responses at any particular time are a random deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse]. As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if these wave-to-wave differences are quite large.

@@

Like with the active updating model, there are theories of attitude development and change at societal level that would require evidence of the settled dispositions model at the individual level. For example, if people's attitudes change in a way that reflects updating, then it would be wrong to consider their deviations from their average to be "measurement error," and theories that suggest that changes in belief are not real [@pageshapiro] would be unsupported. Similarly, an item would have to be relatively stable for most adults for aggregate cultural change to come primarily through cohort replacement [@ryder].

@@

## Consistency in Response

Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change. Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items. As such, we designate a second term, _consistency_, to reflect this second dimension of change.

## Toward Theoretical Synthesis

We do not suggest that all attitudes and behaviors captured by survey questions follow a settled dispositions model or that all display active updating. Instead, we suggest that by classifying the pattern of change at the individual level, we can use that as a tool for adjudicating debates of change at the aggregate level.

The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern. This disagreement thus forms the primary structure of our analysis. However, it is possible that some items display active updating while others display settled dispositions. There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.

_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman]. To put it differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

Items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults but high active updating for young people, as this is the period most commonly found to be susceptible to attitude change [@glenn; @inglehartbaker; @visserkrosnick; @danigelis].

_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate peripheral attitudes, such as views on specific policy questions [@boutylinevaisey; @goldbergstein]. Much work in political science has sought to understand whether people use ideological identity, policy positions, moral beliefs, or partisanship to adjudicate their position on other issues [@lakoff; @greenetal; @kinderkalmoe]. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification as the "core" belief in the political space, consistent with Converse's [-@converse] notion of social constraint. Baldassari and Gelman [-@baldassarigelman] find an increased alignment between partisan identification and policy positions, but little increase in the correlation between specific policy positions over time. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate a highest level.

These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate a cascade of changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs. In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating. Core beliefs should also be more consistently reported, since individuals should know where they stand on these core questions. Determining where one stands on peripheral issues requires greater cognitive effort and therefore are more prone to random perturbations, so should be reported less consistently.

_Public vs. Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances. However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].

Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse]. This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do spotlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, we predict active updating throughout the population, not just young people.

Items that tap some public dimension of behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.

As we noted previously, we do not expect that all items will align with just one model. If we find a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions. Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all “matter.” Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work. We believe achieving a better understanding of these processes will be relevant for many sub-fields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

@@

Third, _are there systematic differences in item content between questions that exhibit different levels of active updating and consistency?_ The preceding sections have made some predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs and behaviors. As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.

@@

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2 (the coefficient on this term should be 0). If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.

@@

That is, the SDM predicts that the average response of previous waves is the best predictor of the next response. If change is non-persisting, then taking the average of the last two responses will be our best guess about a person's underlying position.

@@

### Comparison to Other Approaches
